---
title: "Figure 2 — ABI and NIRS HRT case patient data against normative distributions"
author:
    name: Jem I Arnold [{{< fa brands github >}}](https://github.com/jemarnold/mnirs){target=\"_blank\"} [{{< fa brands linkedin >}}](https://www.linkedin.com/in/jem--arnold/){target=\"_blank\"} [{{< ai researchgate >}}](https://researchgate.net/profile/Jem-Arnold){target=\"_blank\"} [{{< ai orcid >}}](https://orcid.org/0000-0003-3908-9447){target=\"_blank\"}
    email: jem@jemarnold.ca
    affiliation:
        name: |
            School of Kinesiology\
            The University of British Columbia\
            Vancouver, BC, Canada
date: today
format:
  html:
    embed-resources: true
    code-fold: true
    code-tools: true
    toc: true
  gfm:
    output-file: README.md
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.path: "figures/"
    out.width: "100%"
    fig.align: "center"
---

This report reproduces **Figure 2** from the manuscript: *Arnold JI, Pignanelli C, Hodgins A, O'Croinin E, Koehle MS. Return to Sport After Two Sequential Iliac Artery Reconstructions in an Elite Cyclist: A Case Report* (currently pre-submission).

Please contact the authors for further information (author email above).


## Setup

Requires libraries from `tidyverse` and `mnirs` (for plotting only).

```{r}
#| label: setup
library(tibble)
library(tidyr)
library(dplyr)
library(ggplot2)
library(mnirs)

theme_set(theme_mnirs(base_size = 14))

## case report patient data
case_data <- tribble(
    ~trial , ~leg    , ~abi , ~hrt ,
         1 , "left"  , 0.47 ,    54 ,
         2 , "left"  , 0.57 ,    24 ,
         3 , "left"  , 0.59 ,    31 ,
         4 , "left"  , 0.71 ,    31 ,
         5 , "left"  , 0.65 ,    34 ,
         1 , "right" , 0.78 ,    28 ,
         2 , "right" , 0.64 ,    28 ,
         3 , "right" , 0.58 ,    43 ,
         4 , "right" , 0.59 ,    29 ,
         5 , "right" , 0.80 ,    33 ,
) |>
    mutate(
        trial = factor(trial),
        leg = factor(leg, levels = c("left", "right")),
    )
```

## Normative reference statistics

Reference distributions extracted from individual participant data from: van Hooff M, Arnold J, Meijer E, et al. (2022) Diagnosing sport-related flow limitations in the iliac arteries using near-infrared spectroscopy. *J Clin Med*. <https://dx.doi.org/10.3390/jcm11247462>. Raw participant data cannot be shared. The figure below can be reproduced with summary statistics, which are embedded here.

Group sizes (legs with paired ABI and HRT):

| healthy | patient |
|--------:|--------:|
|      58 |     211 |

: {.w-auto}

Derivation from the raw database `df` (one row per leg with `id`, `group`, `leg`, `abi`, `hrt`; not included; not run):

```{r}
#| label: derive
#| eval: false
## robust z-score vs healthy reference: median and IQR-derived SD
## (IQR / 1.349), resistant to outliers and non-normal tails
robust_z <- function(x, ref) {
    return(
        (x - median(ref, na.rm = TRUE)) /
            (IQR(ref, na.rm = TRUE) / (2 * qnorm(0.75)))
    )
}

healthy <- filter(df, group == "Healthy")

## HRT is right-skewed with spread increasing with level (log approximately
## normalises it); ABI is near-symmetric, so it stays on the raw scale
df_z <- df |>
    mutate(
        abi = robust_z(abi, healthy$abi),
        hrt = robust_z(log(hrt), log(healthy$hrt)),
    ) |>
    drop_na(abi, hrt)

## robust centre, spread, and abi-hrt correlation of each group, on the
## healthy z-scale. pearson r of abi vs log(hrt) (z-scaling is linear, so
## r is unchanged)
group_stats <- df_z |>
    summarise(
        .by = group,
        mid_abi = median(abi),
        mid_hrt = median(hrt),
        s_abi = IQR(abi) / (2 * qnorm(0.75)),
        s_hrt = IQR(hrt) / (2 * qnorm(0.75)),
        cor = cor(abi, hrt),
    )
```

Embedded output of the above:

```{r}
#| label: ref-stats
## healthy reference: median and IQR / 1.349 (hrt on log scale)
ref <- tibble(
    abi_centre = 0.716433,
    abi_scale = 0.117873,
    hrt_centre = 2.896581,
    hrt_scale = 0.395201,
)

## group centre, spread, and abi-hrt correlation in healthy z-units. healthy
## centre and spread are 0 / 1 by construction
group_stats <- tribble(
    ~group    , ~mid_abi  , ~mid_hrt , ~s_abi   , ~s_hrt   , ~cor      ,
    "Healthy" ,  0        , 0        , 1        , 1        , -0.244759 ,
    "Patient" , -0.897959 , 2.151128 , 1.043538 , 1.376109 , -0.436214 ,
)

ref
group_stats
```

## Group ellipses and case z-scores

```{r}
#| label: ellipses
## 68% bivariate normal ellipse per group. ABI and HRT paired within leg,
## so ellipse is tilted by their correlation (cholesky of covariance)
ellipses <- group_stats |>
    cross_join(tibble(t = seq(0, 2 * pi, length.out = 200))) |>
    mutate(
        r = sqrt(qchisq(0.68, df = 2)),
        abi = mid_abi + r * s_abi * cos(t),
        hrt = mid_hrt + r * s_hrt * (cor * cos(t) + sqrt(1 - cor^2) * sin(t)),
    )

## case study on same z-scale. HRT log-transformed as for the reference
case_z <- case_data |>
    mutate(
        abi = (abi - ref$abi_centre) / ref$abi_scale,
        hrt = (log(hrt) - ref$hrt_centre) / ref$hrt_scale,
    ) |>
    ## display-only nudge (z-units) to separate overlapping markers and open
    ## space for arrows
    mutate(
        key = paste(leg, trial),
        abi = abi +
            recode_values(
                key,
                "left 3" ~ -0.15,
                "right 4" ~ 0.05,
                default = 0
            ),
        hrt = hrt +
            recode_values(
                key,
                "right 3" ~ 0.08,
                "right 4" ~ -0.08,
                default = 0
            ),
    )

## arrows trimmed at both ends to clear markers. y scaled by panel aspect
## (asp = height / width per z-unit) so gap is uniform on screen
case_arrows <- case_z |>
    arrange(leg, trial) |>
    mutate(
        .by = leg,
        abi_end = lead(abi),
        hrt_end = lead(hrt),
        gap = 0.15,
        asp = 0.6,
        ## per-arrow curvature to route arcs around other markers
        curv = recode_values(
            key,
            "left 2" ~ -0.1,
            "left 3" ~ 0.1,
            "right 1" ~ 0.1,
            "right 4" ~ 0.5,
            default = 0.3
        ),
        len = sqrt((abi_end - abi)^2 + (asp * (hrt_end - hrt))^2),
        dx = gap * (abi_end - abi) / len,
        dy = gap * (hrt_end - hrt) / len,
        abi = abi + dx,
        hrt = hrt + dy,
        abi_end = abi_end - dx,
        hrt_end = hrt_end - dy,
        ## manual arrow offsets (z-units) after auto-trim, keyed by arrow
        ## start point (e.g. "right 4" = arrow 4 -> 5)
        abi = abi +
            recode_values(
                key,
                "left 1" ~ -0.08,
                "left 2" ~ -0.09,
                "left 4" ~ 0.05,
                "right 3" ~ -0.08,
                "right 4" ~ -0.12,
                default = 0
            ),
        hrt = hrt +
            recode_values(
                key,
                "left 1" ~ -0.05,
                "left 2" ~ -0.05,
                "left 4" ~ 0.15,
                "right 4" ~ -0.26,
                default = 0
            ),
        abi_end = abi_end +
            recode_values(
                key,
                "left 1" ~ -0.07,
                "left 2" ~ -0.05,
                "right 4" ~ 0.15,
                default = 0
            ),
        hrt_end = hrt_end +
            recode_values(
                key,
                "left 1" ~ -0.15,
                "left 2" ~ 0.03,
                "left 4" ~ 0.1,
                "right 3" ~ 0.08,
                "right 4" ~ -0.15,
                default = 0
            ),
    ) |>
    drop_na()
```

## Figure

```{r}
#| label: fig-caption
#| echo: false
## shared by html figure chunk and README include chunk
fig_caption <- "Patient ankle-brachial pressure index (ABI) and muscle reoxygenation half-recovery times (HRT), converted to Z-scores based on a cohort of 58 healthy cyclists (green area) and 211 FLIA patients (red area). Shaded clusters represent approximately the 68% distribution of each group. Points numbered 1 to 5 were recorded bilaterally in the patient’s left (pink) and right (blue) legs at each monitoring assessment across the sequential RTS periods. Numbers correspond to the timeline in Figure 1 and values in Table 1. Values closer to the centre of the green “healthy” distribution indicate improvement. The left leg in pink improved from pre-surgery (1) to post-surgery (2), with ABI and HRT maintained until the final assessment (5). The right leg in blue worsened from assessment 1 with symptom onset around assessment 3, then improved again after right-leg surgery in assessments 4 and 5."
```

```{r}
#| label: figure-2
#| fig-width: 8.66
#| fig-height: 5.77
#| fig-dpi: 300
#| fig-cap: !expr fig_caption
# fmt: skip
ggplot() +
    aes(x = abi, y = hrt) +
    labs(x = "ABI (Z-score vs healthy)", y = "HRT (Z-score vs healthy)") +
    coord_cartesian(xlim = c(-3, 3), ylim = c(-2, 4)) +
    scale_colour_manual(
        name = "Patient Leg:",
        aesthetics = c("colour", "fill"),
        values = setNames(
            mnirs::palette_mnirs(
                "light red", "light green", "pink", "light blue"
            ), c("Patient", "Healthy", "left", "right")
        ),
        breaks = c("left", "right")
    ) +
    scale_shape_manual(values = c(left = 21, right = 24)) +
    guides(fill = "none", shape = "none") +
    geom_vline(xintercept = 0, linewidth = 0.5, colour = "grey40") +
    geom_hline(yintercept = 0, linewidth = 0.5, colour = "grey40") +
    geom_polygon(
        data = ellipses,
        aes(colour = group, fill = group),
        linetype = "dashed", linewidth = 0.8, alpha = 0.2, show.legend = FALSE,
    ) +
    geom_label(
        data = transmute(group_stats, group, abi = mid_abi, hrt = mid_hrt),
        aes(label = group, colour = group),
        size = 6, fill = alpha("white", 0.6),
        hjust = c(-0.3, 1.2), vjust = c(2, -3),
        linewidth = NA, fontface = "bold", show.legend = FALSE,
    ) +
    lapply(split(case_arrows, ~curv), \(d) geom_curve(
        data = d,
        aes(xend = abi_end, yend = hrt_end, colour = leg),
        curvature = d$curv[1], linewidth = 1,
        arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
    )) +
    geom_point(
        data = case_z,
        aes(fill = leg, shape = leg),
        size = 9, show.legend = FALSE
    ) +
    geom_text(
        data = case_z,
        aes(label = trial),
        colour = "white", size = 7, fontface = "bold"
    )
```

```{r}
#| label: save-vector
#| eval: !expr knitr::pandoc_to("html")
#| include: !expr knitr::pandoc_to("html")
## high-res of figure above for journal submission (html render only)
ggsave("figures/figure-2.svg", width = 8.66, height = 5.77)
```

## Session info

```{r}
#| label: session
sessionInfo()
```
